---
title: "Project_3_SST"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ncdf4)
```

<< INTRODUCE >>

First, we must begin by loading in the data from the netCDF file format containing the measurements over each month. 

```{r init_load}
# load-in months from data folder (user)
file_1 <- nc_open("../data/matchups_NPP_2018-01.nc")
file_2 <- nc_open("../data/matchups_NPP_2018-02.nc")
file_3 <- nc_open("../data/matchups_NPP_2018-03.nc")
file_4 <- nc_open("../data/matchups_NPP_2018-04.nc")
```

Now, what are the variables available to us within the raw data files?

```{r data_available}
print(file_4)
```

<< MODEL INTRO >>

<< ADD PROBLEM OF ATMOSPHERIC CORRECTION >>

Anding and Kauth (1970) found that the difference in measurement at properly selected infrared (IR) channels is proportional to the required amount of atmospheric correction.

Barton (1995) used this differential absorption between the channels which is used in all IR sea surface temperature (SST) model. In its basic form it usually represented as:

$$
T_S = aT_{\lambda_i}+\gamma(T_{\lambda_i}-T_{\lambda_j})+c
$$

where $T_s$ is estimated SST, $T_{\lambda_i}$ and $T_{\lambda_j}$ are brigthness temperature measurements in channels $\lambda_i$ and $\lambda_j$ where $i\ne j$. Therefore, the trick is to estimated which channels must be used. Both, $a$ and $c$ are constants. Finally, $\gamma$ is defined as

$$
\gamma=\frac{1-\tau_{\lambda_i}}{\tau_{\lambda_i}-\tau_{\lambda_j}}
$$

where $\tau$ is the transmittance through the atmosphere from the sea surface to the signal receiving satellite.

Though all statistical models share the above form, various modifications have been made over time to improve performance. One such model is based on the non-linear SST algorithm (NLSST) which was originally developed by Walton et al. (1998) which has the form:

$$
T_S=b_0+b_1T_{\lambda_i}+b_2(T_{\lambda_i}-T_{\lambda_j})T_{ref}+b_3(T_{\lambda_i}-T_{\lambda_j})(\sec\theta-1)
$$

where $T_{ref}$ is a best-first-guess SST value and $\sec\theta$ is the satellite zenith angle. 


<< PART 1 >>




<< PART 2 >>




<< PART 3 >>




<< PART 4 >>




<< PART 5 >>